---
title: "ST 538 Group 4 Project 2"
format: pdf
---

## Introduction

Vehicle loan default prediction presents a distinct set of challenges relative
to other retail credit products. Unlike mortgage or personal loan portfolios,
two-wheeler financing in emerging markets such as India serves a large
population of first-time borrowers with little or no bureau history, making
traditional score-based underwriting insufficient on its own. This analysis
uses the CRIF High Mark vehicle loan dataset to develop a feature set capable
of identifying first-EMI default risk across both scored and thin-file borrower
segments.

## Data and Processing

The dataset comprises 233,154 disbursed loans with a binary default indicator
capturing payment failure on the first EMI due date. A notable characteristic
of this population is that 56% of borrowers carry no scoreable bureau history,
which necessitates treating thin-file and scored segments with care rather than
imputing or ignoring the distinction. Among scored borrowers, the CRIF CNS
score serves as the single strongest individual predictor and is encoded here
as a 20-level ordered factor preserving the full risk ranking across bands.

Feature engineering focuses on translating domain knowledge into model-ready
variables: loan structure risk (LTV, implied down payment), borrower
demographics (age at disbursement, employment type), identity verification
depth (KYC document count), credit utilisation and delinquency history across
primary and secondary accounts, EMI burden relative to existing obligations,
and recent bureau activity including new account uptake and hard enquiries.

A composite stress score aggregating six binary risk flags is also constructed
as a diagnostic and interpretability aid. Initial cross-tabulation confirms
that the stress score tracks default rate monotonically, rising from 15% at
score 0 to 49% at score 5, providing early evidence that the engineered
features carry meaningful signal.

```{r}
#| include: false
#| warning: false
# load dependencies
library(tidyverse)
library(lubridate)
library(stringr)
library(knitr)
library(randomForest)
```


```{r}
#| include: false
#| echo: false
# Project URL https://www.kaggle.com/datasets/avikpaul4u/vehicle-loan-default-prediction?select=train.csv

set.seed(2026)

# Load our files (saved manually to data directory)
# Training data
train_dat = read.csv('./data/train.csv')

# Testing data (DO NOT LOAD UNTIL WEEK 8 OR 9!!)
#test_dat = read.csv('./data/test.csv')

head(train_dat)
str(train_dat)


```

```{r}
#| include: false
# Check for any columns with missing data
sapply(train_dat, anyNA)
```

There are no missingness in the data! Pretty sweet!




Feature engineering
```{r}
#| include: false
# helper function - parse tenure string to numeric months
parse_tenure_months <- function(x) {
  years  <- as.numeric(str_extract(x, "\\d+(?=yrs)"))
  months <- as.numeric(str_extract(x, "\\d+(?=mon)"))
  years  <- ifelse(is.na(years),  0, years)
  months <- ifelse(is.na(months), 0, months)
  years * 12 + months
}

# loan terms
train_dat <- train_dat %>%
  mutate(
 
    # LTV risk flag — regulatory / underwriting threshold
    ltv_high_flag = as.integer(LTV > 80),
 
    # Implied down payment
    down_payment_amt = ASSET_COST - DISBURSED_AMOUNT,
    down_payment_pct = round((down_payment_amt / ASSET_COST) * 100, 2)
 
  )

# demographics and disbursement date
train_dat <- train_dat %>%
  mutate(
 
    # Parse dates — format is "DD-MM-YYYY" per the raw data
    dob_parsed     = dmy(DATE_OF_BIRTH),
    disbursal_date = dmy(DISBURSAL_DATE),
 
    # Age at disbursement in whole years
    age_at_disbursal = as.integer(
      interval(dob_parsed, disbursal_date) / years(1)
    ),
 
    # Age risk segments
    age_segment = case_when(
      age_at_disbursal <  22                           ~ "young_risky",
      age_at_disbursal >= 22 & age_at_disbursal <= 60  ~ "prime",
      age_at_disbursal >  60                           ~ "senior_risky",
      TRUE                                             ~ NA_character_
    ),
 
    # Disbursement seasonality
    disbursal_month   = month(disbursal_date),
    disbursal_quarter = quarter(disbursal_date),
    disbursal_year    = year(disbursal_date),
 
    # Indian fiscal year-end flag (Jan–Mar = Q4 of Indian FY)
    fy_end_flag = as.integer(disbursal_month %in% c(1, 2, 3))
 
  ) %>%
  select(-dob_parsed)   # intermediate parse column no longer needed

# identity
train_dat <- train_dat %>%
  mutate(
 
    # Total KYC documents submitted (0-6)
    kyc_doc_count = MOBILENO_AVL_FLAG + AADHAR_FLAG + PAN_FLAG +
                    VOTERID_FLAG + DRIVING_FLAG + PASSPORT_FLAG,
 
    # Strong KYC: has both Aadhaar (biometric) AND PAN (income tax records)
    strong_kyc_flag = as.integer(AADHAR_FLAG == 1 & PAN_FLAG == 1)
 
  )

# bureau score

# sort(unique(train_dat$PERFORM_CNS_SCORE_DESCRIPTION)) # check the label strings

# CRIF CNS bands A-M run lowest to highest risk.
# "Not Scored" has 6 distinct reason codes — all treated as thin-file.
# Ordered worst -> best creditworthiness so ordinal models can use the ranking.
 
cns_levels <- c(
  # --- Thin-file / unscoreable (placed at bottom = worst/unknown) ---
  "No Bureau History Available",
  "Not Scored: More than 50 active Accounts found",
  "Not Scored: No Activity seen on the customer (Inactive)",
  "Not Scored: No Updates available in last 36 months",
  "Not Scored: Not Enough Info available on the customer",
  "Not Scored: Only a Guarantor",
  "Not Scored: Sufficient History Not Available",
  # --- Scored bands: highest risk first ---
  "M-Very High Risk",
  "L-Very High Risk",
  "K-High Risk",
  "J-High Risk",
  "I-Medium Risk",
  "H-Medium Risk",
  "G-Low Risk",
  "F-Low Risk",
  "E-Low Risk",
  "D-Very Low Risk",
  "C-Very Low Risk",
  "B-Very Low Risk",
  "A-Very Low Risk"
)
 
# All "Not Scored" reason strings for thin-file flagging
not_scored_reasons <- c(
  "No Bureau History Available",
  "Not Scored: More than 50 active Accounts found",
  "Not Scored: No Activity seen on the customer (Inactive)",
  "Not Scored: No Updates available in last 36 months",
  "Not Scored: Not Enough Info available on the customer",
  "Not Scored: Only a Guarantor",
  "Not Scored: Sufficient History Not Available"
)
 
train_dat <- train_dat %>%
  mutate(
 
    # Ordered factor — preserves risk ranking in tree-based and ordinal models
    cns_score_factor = factor(
      PERFORM_CNS_SCORE_DESCRIPTION,
      levels  = cns_levels,
      ordered = TRUE
    ),
 
    # Thin-file flag: covers all 7 unscoreable categories
    thin_file_flag = as.integer(
      PERFORM_CNS_SCORE_DESCRIPTION %in% not_scored_reasons
    ),
 
    # Thin-file reason — retains the specific not-scored reason for analysis
    # (e.g. "Only a Guarantor" may have a different risk profile than "Inactive")
    thin_file_reason = ifelse(
      thin_file_flag == 1,
      PERFORM_CNS_SCORE_DESCRIPTION,
      NA_character_
    ),
 
    # Clean score: NA out thin-file records so 0 is not treated as a real score
    cns_score_clean = ifelse(thin_file_flag == 1, NA_real_, PERFORM_CNS_SCORE)
    )

# primary accounts
train_dat <- train_dat %>%
  mutate(
 
    # Active loan utilisation ratio
    pri_active_util_ratio = ifelse(
      PRI_NO_OF_ACCTS > 0,
      round(PRI_ACTIVE_ACCTS / PRI_NO_OF_ACCTS, 4),
      0
    ),
 
    # Any overdue primary account — binary red flag
    pri_any_overdue_flag = as.integer(PRI_OVERDUE_ACCTS > 0),
 
    # Incremental leverage: existing balance relative to new loan size
    pri_leverage_ratio = ifelse(
      DISBURSED_AMOUNT > 0,
      round(PRI_CURRENT_BALANCE / DISBURSED_AMOUNT, 4),
      NA_real_
    ),
 
    # Average primary loan size historically
    pri_avg_loan_size = ifelse(
      PRI_NO_OF_ACCTS > 0,
      round(PRI_DISBURSED_AMOUNT / PRI_NO_OF_ACCTS, 2),
      0
    ),
 
    # Sanctioned vs. disbursed gap (large gap may indicate cancellations)
    pri_sanction_disbursed_gap = PRI_SANCTIONED_AMOUNT - PRI_DISBURSED_AMOUNT
 
  )

# joint accounts
train_dat <- train_dat %>%
  mutate(
 
    # Binary: any secondary overdue
    sec_any_overdue_flag  = as.integer(SEC_OVERDUE_ACCTS > 0),
 
    # Has any secondary exposure at all
    sec_has_exposure_flag = as.integer(SEC_NO_OF_ACCTS > 0)
 
  )

# EMI / debt service
train_dat <- train_dat %>%
  mutate(
 
    # Total monthly EMI obligation across primary and secondary loans
    total_emi_burden = PRIMARY_INSTAL_AMT + SEC_INSTAL_AMT,
 
    # New loan EMI relative to existing primary EMI commitment
    # Ratio > 1: new loan eclipses all current EMI obligations combined
    new_loan_emi_ratio = ifelse(
      PRIMARY_INSTAL_AMT > 0,
      round(DISBURSED_AMOUNT / PRIMARY_INSTAL_AMT, 4),
      NA_real_
    )
 
  )

# recent bureau activity
train_dat <- train_dat %>%
  mutate(
 
    # Binary: any delinquency in the last 6 months
    recent_delinquency_flag = as.integer(DELINQUENT_ACCTS_IN_LAST_SIX_MONTHS > 0),
 
    # Inquiry-to-new-account ratio
    # When NEW_ACCTS = 0: return raw inquiry count — all enquiries were rejected
    inquiry_to_new_acct_ratio = ifelse(
      NEW_ACCTS_IN_LAST_SIX_MONTHS > 0,
      round(NO_OF_INQUIRIES / NEW_ACCTS_IN_LAST_SIX_MONTHS, 4),
      NO_OF_INQUIRIES
    ),
 
    # Parse "Xyrs Ymon" tenure strings to numeric months
    avg_acct_age_months   = parse_tenure_months(AVERAGE_ACCT_AGE),
    credit_history_months = parse_tenure_months(CREDIT_HISTORY_LENGTH),
 
    # Thin credit history flag (<12 months = thin-file risk)
    short_history_flag = as.integer(credit_history_months < 12)
 
  )

# composite financial stress score
train_dat <- train_dat %>%
  mutate(
 
    # Additive count of binary risk flags
    # Useful for EDA, stakeholder comms, and model sanity checks
    stress_score = pri_any_overdue_flag    +
                   sec_any_overdue_flag    +
                   recent_delinquency_flag +
                   ltv_high_flag           +
                   short_history_flag      +
                   thin_file_flag,
 
    # High-stress flag: 3 or more simultaneous risk indicators
    high_stress_flag = as.integer(stress_score >= 3)
 
  )

# employment type na results fix
train_dat <- train_dat %>%
    mutate(EMPLOYMENT_TYPE = na_if(EMPLOYMENT_TYPE, ""))
```





```{r}
# validation
cat("Rows:", nrow(train_dat), "\n")
cat("Total columns after engineering:", ncol(train_dat), "\n")
cat("Net new columns:", ncol(train_dat) - 41, "\n\n")
 
cat("--- LTV high flag (>80%) ---\n")
print(table(train_dat$ltv_high_flag, useNA = "ifany"))
 
cat("\n--- Age segment ---\n")
print(table(train_dat$age_segment, useNA = "ifany"))
 
cat("\n--- Employment type ---\n")
print(table(train_dat$EMPLOYMENT_TYPE, useNA = "ifany"))
 
cat("\n--- CNS score factor levels (confirm all levels matched) ---\n")
print(levels(train_dat$cns_score_factor))
cat("Unmatched descriptions (NAs in factor):",
    sum(is.na(train_dat$cns_score_factor)), "\n")
 
cat("\n--- Thin-file borrowers ---\n")
print(table(train_dat$thin_file_flag, useNA = "ifany"))
 
cat("\n--- Primary any overdue ---\n")
print(table(train_dat$pri_any_overdue_flag, useNA = "ifany"))
 
cat("\n--- Recent delinquency flag ---\n")
print(table(train_dat$recent_delinquency_flag, useNA = "ifany"))
 
cat("\n--- Stress score distribution ---\n")
print(table(train_dat$stress_score, useNA = "ifany"))
 
cat("\n--- avg_acct_age_months (first 10 rows) ---\n")
print(data.frame(
  raw    = head(train_dat$AVERAGE_ACCT_AGE, 10),
  parsed = head(train_dat$avg_acct_age_months, 10)
))
 
cat("\n--- credit_history_months (first 10 rows) ---\n")
print(data.frame(
  raw    = head(train_dat$CREDIT_HISTORY_LENGTH, 10),
  parsed = head(train_dat$credit_history_months, 10)
))
```



```{r}
# default rate by stress score table
stress_table <- train_dat %>%
  group_by(stress_score) %>%
  summarise(
    n           = n(),
    n_default   = sum(LOAN_DEFAULT, na.rm = TRUE),
    .groups     = "drop"
  ) %>%
  mutate(
    n_no_default  = n - n_default,
    default_rate  = paste0(round(n_default / n * 100, 1), "%")
  ) %>%
  select(
    `Stress score`  = stress_score,
    `Total (n)`     = n,
    `No default`    = n_no_default,
    `Default`       = n_default,
    `Default rate`  = default_rate
  )

knitr::kable(
  stress_table,
  format      = "simple",
  caption     = "Default Rate by Stress Score",
  align       = c("c", "r", "r", "r", "r"),
  format.args = list(big.mark = ",")
)
```



```{r}
# default rate by stress score graphically
stress_summary <- train_dat %>%
  group_by(stress_score) %>%
  summarise(
    n            = n(),
    n_default    = sum(LOAN_DEFAULT, na.rm = TRUE),
    .groups      = "drop"
  ) %>%
  mutate(default_rate = n_default / n * 100)
 
scale_factor <- max(stress_summary$n) / 50
 
ggplot(stress_summary, aes(x = factor(stress_score))) +
 
  geom_col(aes(y = n), width = 0.6) +
 
  geom_line(aes(y = default_rate * scale_factor, group = 1)) +
  geom_point(aes(y = default_rate * scale_factor)) +
 
  geom_text(
    aes(y = default_rate * scale_factor, label = paste0(round(default_rate, 1), "%")),
    vjust = -0.9, size = 3.2
  ) +
 
  scale_y_continuous(
    name     = "Count (n)",
    labels   = scales::comma,
    sec.axis = sec_axis(~ . / scale_factor, name = "Default rate (%)",
                        labels = scales::label_number(suffix = "%"))
  ) +
 
  theme_classic(base_size = 12) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank()) +
 
  labs(
    title = "Default rate rises with stress score",
    x     = "Stress score"
  )
```

Summary of EDA and feature engineering:
Thirty-five percent of loans are high loan to value (LTV) of >80%, combined with a large portion of borrowers having thin credit history (%), it shows a meaningful concentration of risk. The vast majority (94%) of borrowers fall in the prime age band, 13,812 young borrowers (<22 years of age) are worth watching. Fifty-six percent of borrowers are "thin file" meaning they have <12 month of credit history. This makes sense, in India about half of first time car buyers are using lending for the first time. Thin and normal are two different populations, we may want to split the dataset and create a model for thin credit and one for normal, or focus our efforts on normal only--the model will likely suffer trying to work on both. 

Stress score is working well as a risk signal. Default raises from 15% at score 0 to ~49% at score 5 (3x). Most of the population sits at 2 (21% default), probably thin-file + high-LTV, but no active delinquncy.The jump from 2 to 3 is steep (21-29%), pri_any_overdue_flag and/or recent_delinquncy_flag are likely doing a lot of work. Scores 4 and 5 are small segments with high default rates (42-49%), it's really up to us if we want to keep them, probably should. Score 0 is not zero (15.1%), even the cleanest borrowers carry default risk. We can use this stress_score as a standalone feature, but it's worth noting that components might actually capture more non-linear interactions than the score itself.




```{r}
# engineered feature index table
feature_index <- tribble(
  ~Category,          ~Feature,
  "Loan terms",       "ltv_high_flag",
  "Loan terms",       "down_payment_amt",
  "Loan terms",       "down_payment_pct",
  "Demographics",     "age_at_disbursal",
  "Demographics",     "age_segment",
  "Demographics",     "disbursal_month",
  "Demographics",     "disbursal_quarter",
  "Demographics",     "disbursal_year",
  "Demographics",     "fy_end_flag",
  "KYC",              "kyc_doc_count",
  "KYC",              "strong_kyc_flag",
  "Bureau score",     "cns_score_factor",
  "Bureau score",     "thin_file_flag",
  "Bureau score",     "thin_file_reason",
  "Bureau score",     "cns_score_clean",
  "Primary credit",   "pri_active_util_ratio",
  "Primary credit",   "pri_any_overdue_flag",
  "Primary credit",   "pri_leverage_ratio",
  "Primary credit",   "pri_avg_loan_size",
  "Primary credit",   "pri_sanction_disbursed_gap",
  "Secondary credit", "sec_any_overdue_flag",
  "Secondary credit", "sec_has_exposure_flag",
  "EMI / DSCR",       "total_emi_burden",
  "EMI / DSCR",       "new_loan_emi_ratio",
  "Recent activity",  "recent_delinquency_flag",
  "Recent activity",  "inquiry_to_new_acct_ratio",
  "Recent activity",  "avg_acct_age_months",
  "Recent activity",  "credit_history_months",
  "Recent activity",  "short_history_flag",
  "Composite",        "stress_score",
  "Composite",        "high_stress_flag"
)
 
# collapse each category's features into a single comma-separated string
feature_index_wide <- feature_index %>%
  group_by(Category) %>%
  summarise(Features = paste(Feature, collapse = ", "), .groups = "drop")
 
knitr::kable(feature_index_wide, format = "simple", col.names = c("Category", "Features"),
             caption = "Feature Index Table")
```




```{r}
train_dat_normal <- train_dat %>%
  filter(thin_file_flag == 0)

train_dat_thin <- train_dat %>%
  filter(thin_file_flag == 1)

cbind(count(train_dat_normal),count(train_dat_thin))
```






```{r}
#| include: false

# Set our splits
train_percent <- 0.7
not_train_percent <- 0.3

# not_train split up 
validate_percent <- 0.5 # Ends up being 0.15
train_percent <- 0.5 # Ends up being 0.15


# splitting into train/not_train -   this had an error with nrow(train_dat) producing too long of a sample
sample <- sample(c(TRUE, FALSE), nrow(train_dat_normal), replace=TRUE, prob=c(train_percent,not_train_percent))
training  <- train_dat_normal[sample, ]



not_training   <- train_dat_normal[!sample, ]

# splitting into validation/test
train_sample <- sample(c(TRUE, FALSE), nrow(not_training), replace=TRUE, prob=c(validate_percent,train_percent))
validation  <- not_training[train_sample, ]
testing   <- not_training[!train_sample, ]


# # Did they default?
# train_dat |>
#   summarise(
#     defaulted_yes_overall = sum(LOAN_DEFAULT == 1),
#     defaulted_no_overall = sum(LOAN_DEFAULT == 0)
#     
#   )
# 
# train |>
#   summarise(
#     defaulted_yes_train = sum(LOAN_DEFAULT == 1),
#     defaulted_no_train = sum(LOAN_DEFAULT == 0)
#   )
# 
# test |>
#   summarise(
#     defaulted_yes_test = sum(LOAN_DEFAULT == 1),
#     defaulted_no_test = sum(LOAN_DEFAULT == 0)
#     
#   )
```


```{r}
  #sapply(testing[,c(-10, -56, -66, -73)], anyNA)
```

```{r}
  head(training[,c(-10, -56, -66, -73)])
  # the following have NA values which do not work with forest
  #10 EMPLOYMENT_TYPE
  #56 thin_file_reason
  #66 new_loan_emi_ratio
  
  #73 is predictor high_stress_flag
  #sapply(training[,c(-10, -56, -66, -73)], anyNA)
  
  #training |> filter(is.na(thin_file_reason))
  #training |> filter(is.na(high_stress_flag))
```


```{r}
  rf.sim1 = randomForest(training[,c(-10, -56, -66, -73)], 
                         factor(training[,73]))
```


```{r}
# CURRENTLY TOO LARGE!!!
# Random Forest Classifier for Digit Recognition Data
  #rf.digit = randomForest(training[,c(-10, -56, -66, -73)], factor(training[,73]),  xtest = testing[,c(-10, -56, -66, -73)], factor(testing[,73]), ntrees = 7,  importance = T, proximity = T)
  # Predicted response labels
  #rfPred.digit = (rf.digit$test)$predicted
  # Misclassification errors in the test data
 # table(digit_no.test, rfPred.digit)  
  # Variable Importance Plot
  #varImpPlot(rf.digit)  
  

```

```{r}
  pred.rf.sim1 = rf.sim1$predicted
  table(training[,73], pred.rf.sim1)
  
  varImpPlot(rf.sim1)  
  
  
  #training |> filter(high_stress_flag == 1)
```


```{r}
  
```

